Component response rate variation drives stability in large complex systems

### 2018-05-22 ms.Rmd
Restructure manuscript

---
title: "Component response rate variation drives stability in large complex systems"
author: "Brad Duthie"
bibliography: references.bib
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

Rows in $M$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a scalar value $\gamma_{i}$[@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $M$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $M$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $y_{i}$ to generate $M$. When $\gamma_{i} = 1$, eigenvalues of $M$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (grey circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$) and all $\Re({\lambda}) < 0$. The latter cluster has real components that have shifted to the right, but has a smaller radius. Overall, for 1 million randomly assembled $M$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,2])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

@@

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina]. These include systems in which all interactions (i.e., all off-diagonal elements of $M$) are negative (e.g., competitive networks), positive (e.g., mutualist networks), or $i$ and $j$ pairs have opposing signs (e.g., predator-prey networks). In general, competitive and mutualist networks tend to be stabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist communities does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate[@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

@@

**Figure 3: Stability of large complex systems with and without variation in component response rate ($\boldsymbol{\gamma}$).** The $\ln$ number of systems that are stable across different system sizes ($S$) given $C = 1$, and the proportion of systems in which variation in $\gamma$ is critical for system stability. For each $S$, 1 million complex systems are randomly generated. Stability of each complex system is tested given variation in $\gamma$ by randomly sampling $\gamma \sim \mathcal{U}(0, 2)$. Stability given $Var(\gamma)$ is then compared to stability in an otherwise identical system in which $\gamma = E[\mathcal{U}(0, 2)]$ for all components. Light and dark grey bars show the number of stable systems in the absence and presence of variance in $\gamma$, respectively. The black line shows the proportion of systems that are stable when $Var(\gamma) > 0$, but would be unstable if $Var(\gamma) = 0$.

**Figure 4: Distributions of variation in component response time evolved to generate stable systems**

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "Brad Duthie"
bibliography: references.bib
output:
  html_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

Rows in $M$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a scalar value $\gamma_{i}$[@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $M$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $M$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $y_{i}$ to generate $M$. When $\gamma_{i} = 1$, eigenvalues of $M$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (grey circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$) and all $\Re({\lambda}) < 0$. The latter cluster has real components that have shifted to the right, but has a smaller radius. Overall, for 1 million randomly assembled $M$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,1])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

**Figure 3: Stability of large complex systems with and without variation in component response rate($\boldsymbol{\gamma}$).** The $\ln$ number of systems that are stable across different system sizes ($S$) given $C = 1$, and the proportion of systems in which variation in $\gamma$ is critical for system stability. For each $S$, 1 million complex systems are randomly generated. Stability of each complex system is tested given variation in $\gamma$ by randomly sampling $\gamma \sim \mathcal{U}(0, 2)$. Stability given $Var(\gamma)$ is then compared to stability in an otherwise identical system in which $\gamma = E[\mathcal{U}(0, 2)]$ for all components. Light and dark grey bars show the number of stable systems in the absence and presence of variance in $\gamma$, respectively. The black line shows the proportion of systems that are stable when $Var(\gamma) > 0$, but would be unstable if $Var(\gamma) = 0$.

@@

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina2012]. These include systems in which all interactions (i.e., all off-diagonal elements of $M$) are negative (e.g., competitive networks), positive (e.g., mutualist networks), or $i$ and $j$ pairs have opposing signs (e.g., predator-prey networks). In general, competitive and mutualist networks tend to be stabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist communities does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate[@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

@@

### 2018-05-22 ms.tex
Restructure manuscript

Rows in \(M\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a scalar value
\(\gamma_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{14}}.
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(M\) where \(\gamma_{i} = 1\) for all \(i\) in \(S\) to
the same \(M\) when half of \(\gamma_{i} = 1.95\) and half of
\(\gamma_{i} = 0.05\). This models one system in which \(\gamma\) is
invariant and one in which \(\gamma\) varies, but systems are otherwise
identical (note \(E[\gamma_{i}] = 1\) in both cases). I assume
\(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal elements are
set to \(-1\) and non-zero off-diagonal elements are drawn from
\(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied by \(y_{i}\) to
generate \(M\). When \(\gamma_{i} = 1\), eigenvalues of \(M\) are
distributed uniformly within a circle centred at (\(-1, 0\)) with a
radius of 1.265 (Fig. 1a). Hence, the real components of eigenvalues are
highly unlikely to all be negative when all \(\gamma_{i} = 1\). But when
\(\gamma_{i}\) values are separated into two groups, eigenvalues are no
longer uniformly distributed (Fig. 1b). Instead, two distinct clusters
of eigenvalues appear (grey circles in Fig. 1b), one centred at
(\(-1.95, 0\)) and the other centred at (\(-0.05, 0\)). The former has a
large radius, but the real components have shifted to the left (in
comparison to when \(\gamma = 1\)) and all \(\Re({\lambda}) < 0\). The
latter cluster has real components that have shifted to the right, but
has a smaller radius. Overall, for 1 million randomly assembled \(M\),
this division between slow and fast component response rates results in
more stable systems: 0 stable given \(\gamma = 1\) versus 0 stable given
\(\gamma = \{1.95, 0.5\}\).

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I now
focus on a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\)
(see Supporting Information for other distributions, which give similar
results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig. 1b),
\(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(M\) before and after variation in component response
rate. Figure 2 shows a comparison of eigenvalue distributions given
\(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{17}}, when
\(\gamma = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 12.649.
Uniform variation in \(\gamma\) leads to a non-uniform distribution of
eigenvalues, some of which are clustered tightly around the centre of
the distribution, but others of which are spread outside the former
radius of 12.649 (red circle Fig 2b). This larger radius occurs because
the addition of \(Var(\gamma)\) increases the realised \(\sigma\) of
\(M\). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold of \(\sigma\sqrt{SC} > 1\).

To investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from \(2-32\) (see Supporting Information for different
values of \(\sigma\) and \(C\)). One million \(M\) were simulated for
each \(S\), and the stability of \(M\) was assessed given \(\gamma = 1\)
versus \(\gamma \sim \mathcal{U}(0, 2)\) (note that under these
conditions, \(\sigma\sqrt{SC} = 1\) given \(S = 25\) when
\(\gamma = 1\)). I found that the number of stable random systems was
consistently higher given \(Var(\gamma)\) than when \(\gamma = 1\) (Fig.
3), and that the difference between the probabilities of observing a
stable system increased with an increase in \(S\); i.e., the potential
for \(Var(\gamma)\) to drive stability increased with system complexity.
For the highest values of \(S\), nearly all systems that were stable
given \(Var(\gamma)\) would not have been stable given \(\gamma = 1\),
and the maximum observed \(S\) for which a system was stable was \(31\)
given \(Var(\gamma)\) versus \(27\) given \(\gamma = 1\) (see Supporting
Information for full results). This suggests that the stability of large
systems might be dependent upon variation in the response rate of their
individual components, meaning that factors such as generation time (in
ecological networks), transaction speed (in economic networks), or
communication speed (in social networks) needs to be considered when
investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise
that \(Var(\gamma)\) is not stabilising per se; that is, adding
variation in \(\gamma\) to a particular system \(M\) does not
necessarily increase the probability that the system will be stable (see
Supporting Information). Rather, systems that are observed to be stable
are more likely to vary in \(\gamma\), and for this \(Var(\gamma)\) to
be critical to their stability. This is caused by the shift in the
distribution of eigenvalues that occurs by introducing \(Var(\gamma)\)
(Fig. 1b, 2b), which can sometimes result in all \(\Re(\lambda) < 0\)
but might also increase \(\Re(\lambda)\) values.

@@

\textbf{Figure 1: Example distribution of eigenvalues before (a) and
after (b) separating a randomly generated complex system into fast
(\(\boldsymbol{\gamma} = 1.95\)) and slow
(\(\boldsymbol{\gamma} = 0.05\)) component response rates.} Each panel
shows the same system where \(S = 200\), \(C = 0.05\), and
\(\sigma = 0.4\), and in each case \(E[\gamma] = 1\) (i.e., only the
distribution of \(\gamma\) differs between panels). \textbf{a.}
Eigenvalues plotted when all \(\gamma = 1\); distributions of points are
uniformly distributed within the grey circle with a radius of
\(\sigma\sqrt{SC} =\) 1.238 centred at -1 on the real axis. \textbf{b.}
Eigenvalues plotted when half \(\gamma = 1.95\) and half
\(\gamma = 0.05\); distributions of points can be partitioned into one
large circle of \(\sigma\sqrt{SC} =\) 1.718 centred at
\(\gamma = -1.95\) and one small circle of \(\sigma\sqrt{SC} =\) 0.044
centred at \(\gamma = -0.05\). In a, the maximum real eigenvalaue
\(\max\left(\Re(\lambda)\right) =\) 0.2344871, while in b
\(\max\left(\Re(\lambda)\right) =\) -0.0002273135, meaning that the
complex system in b but not a is stable because in b
\(\max\left(\Re(\lambda)\right) < 0\). In 1 million randomly generated
complex systems under the same parameter values, 0 were stable when
\(\gamma = 1\) while 0 were stable when \(\gamma = \{1.95, 0.05\}\).
Overall, complex systems that are separated into fast versus slow
components tend to be more stable than otherwise identical systems with
identical component response rates.

\includegraphics{ms_files/figure-latex/unnamed-chunk-6-1.pdf}

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.555 and 0.549 for a and b, respectively.

\includegraphics{ms_files/figure-latex/unnamed-chunk-8-1.pdf}

\textbf{Figure 3: Stability of large complex systems with and without
variation in component response rate (\(\boldsymbol{\gamma}\)).} The
\(\ln\) number of systems that are stable across different system sizes
(\(S\)) given \(C = 1\), and the proportion of systems in which
variation in \(\gamma\) is critical for system stability. For each
\(S\), 1 million complex systems are randomly generated. Stability of
each complex system is tested given variation in \(\gamma\) by randomly
sampling \(\gamma \sim \mathcal{U}(0, 2)\). Stability given
\(Var(\gamma)\) is then compared to stability in an otherwise identical
system in which \(\gamma = E[\mathcal{U}(0, 2)]\) for all components.
Light and dark grey bars show the number of stable systems in the
absence and presence of variance in \(\gamma\), respectively. The black
line shows the proportion of systems that are stable when
\(Var(\gamma) > 0\), but would be unstable if \(Var(\gamma) = 0\).

\begin{verbatim}
## Warning in strwidth(legend, units = "user", cex = cex, font = text.font):
## conversion failure on 'γ = 1' in 'mbcsToSbcs': dot substituted for <ce>
\end{verbatim}

\begin{verbatim}
## Warning in strwidth(legend, units = "user", cex = cex, font = text.font):
## conversion failure on 'γ = 1' in 'mbcsToSbcs': dot substituted for <b3>
\end{verbatim}

\begin{verbatim}
## Warning in strwidth(legend, units = "user", cex = cex, font = text.font):
## conversion failure on 'Var(γ)' in 'mbcsToSbcs': dot substituted for <ce>
\end{verbatim}

\begin{verbatim}
## Warning in strwidth(legend, units = "user", cex = cex, font = text.font):
## conversion failure on 'Var(γ)' in 'mbcsToSbcs': dot substituted for <b3>
\end{verbatim}

\begin{verbatim}
## Warning in text.default(x, y, ...): conversion failure on 'γ = 1' in
## 'mbcsToSbcs': dot substituted for <ce>
\end{verbatim}

\begin{verbatim}
## Warning in text.default(x, y, ...): conversion failure on 'γ = 1' in
## 'mbcsToSbcs': dot substituted for <b3>
\end{verbatim}

\begin{verbatim}
## Warning in text.default(x, y, ...): conversion failure on 'Var(γ)' in
## 'mbcsToSbcs': dot substituted for <ce>
\end{verbatim}

\begin{verbatim}
## Warning in text.default(x, y, ...): conversion failure on 'Var(γ)' in
## 'mbcsToSbcs': dot substituted for <b3>
\end{verbatim}

\textbf{Figure 4: Distributions of variation in component response time
evolved to generate stable systems}

>>>>

\textbf{Figure 1: Example distribution of eigenvalues before (a) and
after (b) separating a randomly generated complex system into fast
(\(\boldsymbol{\gamma} = 1.95\)) and slow
(\(\boldsymbol{\gamma} = 0.05\)) component response rates.} Each panel
shows the same system where \(S = 200\), \(C = 0.05\), and
\(\sigma = 0.4\), and in each case \(E[\gamma] = 1\) (i.e., only the
distribution of \(\gamma\) differs between panels). \textbf{a.}
Eigenvalues plotted when all \(\gamma = 1\); distributions of points are
uniformly distributed within the grey circle with a radius of
\(\sigma\sqrt{SC} =\) 1.238 centred at -1 on the real axis. \textbf{b.}
Eigenvalues plotted when half \(\gamma = 1.95\) and half
\(\gamma = 0.05\); distributions of points can be partitioned into one
large circle of \(\sigma\sqrt{SC} =\) 1.718 centred at
\(\gamma = -1.95\) and one small circle of \(\sigma\sqrt{SC} =\) 0.044
centred at \(\gamma = -0.05\). In a, the maximum real eigenvalaue
\(\max\left(\Re(\lambda)\right) =\) 0.2344871, while in b
\(\max\left(\Re(\lambda)\right) =\) -0.0002273135, meaning that the
complex system in b but not a is stable because in b
\(\max\left(\Re(\lambda)\right) < 0\). In 1 million randomly generated
complex systems under the same parameter values, 1 were stable when
\(\gamma = 1\) while 32 were stable when \(\gamma = \{1.95, 0.05\}\).
Overall, complex systems that are separated into fast versus slow
components tend to be more stable than otherwise identical systems with
identical component response rates.

\includegraphics{ms_files/figure-latex/unnamed-chunk-4-1.pdf}

Rows in \(M\) define how a given component \(i\) is affected by other
components of the system, meaning that the rate of component response
time can be modelled by multiplying all row elements by a scalar value
\(\gamma_{i}\)\textsuperscript{\protect\hyperlink{ref-Patel2018}{14}}.
The distribution of \(\gamma\) over \(S\) components thereby models the
distribution of component response rates. An instructive example
compares one \(M\) where \(\gamma_{i} = 1\) for all \(i\) in \(S\) to
the same \(M\) when half of \(\gamma_{i} = 1.95\) and half of
\(\gamma_{i} = 0.05\). This models one system in which \(\gamma\) is
invariant and one in which \(\gamma\) varies, but systems are otherwise
identical (note \(E[\gamma_{i}] = 1\) in both cases). I assume
\(S = 200\), \(C = 0.05\), and \(\sigma = 0.4\); diagonal elements are
set to \(-1\) and non-zero off-diagonal elements are drawn from
\(\mathcal{N}(0, \sigma^{2})\). Rows are then multiplied by \(y_{i}\) to
generate \(M\). When \(\gamma_{i} = 1\), eigenvalues of \(M\) are
distributed uniformly within a circle centred at (\(-1, 0\)) with a
radius of 1.265 (Fig. 1a). Hence, the real components of eigenvalues are
highly unlikely to all be negative when all \(\gamma_{i} = 1\). But when
\(\gamma_{i}\) values are separated into two groups, eigenvalues are no
longer uniformly distributed (Fig. 1b). Instead, two distinct clusters
of eigenvalues appear (grey circles in Fig. 1b), one centred at
(\(-1.95, 0\)) and the other centred at (\(-0.05, 0\)). The former has a
large radius, but the real components have shifted to the left (in
comparison to when \(\gamma = 1\)) and all \(\Re({\lambda}) < 0\). The
latter cluster has real components that have shifted to the right, but
has a smaller radius. Overall, for 1 million randomly assembled \(M\),
this division between slow and fast component response rates results in
more stable systems: 1 stable given \(\gamma = 1\) versus 32 stable
given \(\gamma = \{1.95, 0.5\}\).

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.549 and 0.553 for a and b, respectively.

\includegraphics{ms_files/figure-latex/unnamed-chunk-7-1.pdf}

Higher stability in systems with variation in \(\gamma\) can be observed
by sampling \(\gamma_{i}\) values from various distributions. I now
focus on a uniform distribution where \(\gamma \sim \mathcal{U}(0, 2)\)
(see Supplementary Information for other distributions, which give
similar results). As with the case of \(\gamma = \{1.95, 0.5\}\) (Fig.
1b), \(E[\gamma] = 1\) when \(\gamma \sim \mathcal{U}(0, 2)\), allowing
comparison of \(M\) before and after variation in component response
rate. Figure 2 shows a comparison of eigenvalue distributions given
\(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). As
expected\textsuperscript{\protect\hyperlink{ref-Tao2010}{17}}, when
\(\gamma = 1\), eigenvalues are distributed uniformly in a circle
centred at (\(-1, 0\)) with a radius of \(\sigma\sqrt{SC} =\) 12.649.
Uniform variation in \(\gamma\) leads to a non-uniform distribution of
eigenvalues, some of which are clustered tightly around the centre of
the distribution, but others of which are spread outside the former
radius of 12.649 (red circle Fig 2b). This larger radius occurs because
the addition of \(Var(\gamma)\) increases the realised \(\sigma\) of
\(M\). The clustering and spreading of eigenvalues introduced by
\(Var(\gamma)\) can destabilise previously stable systems or stabilise
systems that are otherwise unstable. But where systems are otherwise too
complex to be stable given \(\gamma = 1\), the effect of \(Var(\gamma)\)
can often lead to stability above
May's\textsuperscript{\protect\hyperlink{ref-May1972}{1},\protect\hyperlink{ref-Allesina2012}{2}}
threshold of \(\sigma\sqrt{SC} > 1\).

\textbf{Figure 3: Stability of large complex systems with and without
variation in component response rate(\(\boldsymbol{\gamma}\)).} The
\(\ln\) number of systems that are stable across different system sizes
(\(S\)) given \(C = 1\), and the proportion of systems in which
variation in \(\gamma\) is critical for system stability. For each
\(S\), 1 million complex systems are randomly generated. Stability of
each complex system is tested given variation in \(\gamma\) by randomly
sampling \(\gamma \sim \mathcal{U}(0, 2)\). Stability given
\(Var(\gamma)\) is then compared to stability in an otherwise identical
system in which \(\gamma = E[\mathcal{U}(0, 2)]\) for all components.
Light and dark grey bars show the number of stable systems in the
absence and presence of variance in \(\gamma\), respectively. The black
line shows the proportion of systems that are stable when
\(Var(\gamma) > 0\), but would be unstable if \(Var(\gamma) = 0\).

To investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from \(2-32\) (see Supplementary Information for different
values of \(\sigma\) and \(C\)). One million \(M\) were simulated for
each \(S\), and the stability of \(M\) was assessed given \(\gamma = 1\)
versus \(\gamma \sim \mathcal{U}(0, 2)\) (note that under these
conditions, \(\sigma\sqrt{SC} = 1\) given \(S = 25\) when
\(\gamma = 1\)). I found that the number of stable random systems was
consistently higher given \(Var(\gamma)\) than when \(\gamma = 1\) (Fig.
3), and that the difference between the probabilities of observing a
stable system increased with an increase in \(S\); i.e., the potential
for \(Var(\gamma)\) to drive stability increased with system complexity.
For the highest values of \(S\), nearly all systems that were stable
given \(Var(\gamma)\) would not have been stable given \(\gamma = 1\),
and the maximum observed \(S\) for which a system was stable was \(31\)
given \(Var(\gamma)\) versus \(27\) given \(\gamma = 1\) (see
Supplementary Information for full results). This suggests that the
stability of large systems might be dependent upon variation in the
response rate of their individual components, meaning that factors such
as generation time (in ecological networks), transaction speed (in
economic networks), or communication speed (in social networks) needs to
be considered when investigating the stability of complex systems.

Some care is needed when interpreting these results. First, I emphasise
that \(Var(\gamma)\) is not stabilising per se; that is, adding
variation in \(\gamma\) to a particular system \(M\) does not
necessarily increase the probability that the system will be stable (see
Supplementary Information). Rather, systems that are observed to be
stable are more likely to vary in \(\gamma\), and for this
\(Var(\gamma)\) to be critical to their stability. This is caused by the
shift in the distribution of eigenvalues that occurs by introducing
\(Var(\gamma)\) (Fig. 1b, 2b), which can sometimes result in all
\(\Re(\lambda) < 0\) but might also increase \(\Re(\lambda)\) values. To
further investigate the potential of \(Var(\gamma)\) to be stabilising,
I used a simple genetic algorithm because the space of possible
\(\gamma\) values was too large to searh
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{18}}
(see Supplementary Information). For each of 10000 random \(M\), the
genetic algorithm initialised 1000 different sets of
\(\gamma \sim \mathcal{U}(0, 2)\) values of size \(S\). Eigenanalysis of
each set of \(\gamma\) values was performed, and the 20 sets with the
lowest \(\max\left(\Re(\lambda)\right)\) each produced 50 offspring with
subsequent mutation and crossover between the resulting new population
of 1000 \(\gamma\) sets. The genetic algorithm terminated if a stable
\(M\) was found, 20 generations occurred, or a converge criteria of
minimum fitness increase was satisfied. Across
\(S = \{2, 3, ..., 49, 50\}\), sets of \(\gamma\) values were frequently
found result in stable systems, suggesting that varying component
response time might by itself be sufficient to stabilise complex
systems.

I have focused broadly on random complex systems, but it is also
worthwhile to consider more restricted interactions such as those of
specific ecological
networks\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. These
include systems in which all interactions (i.e., all off-diagonal
elements of \(M\)) are negative (e.g., competitive networks), positive
(e.g., mutualist networks), or \(i\) and \(j\) pairs have opposing signs
(e.g., predator-prey networks). In general, competitive and mutualist
networks tend to be stabilising, and predator-prey network tend to be
stabilising\textsuperscript{\protect\hyperlink{ref-Allesina2011}{19}}.
When \(Var(\gamma)\) is applied to each, the proportion of stable
competitive and predator-prey networks increases, but the proportion of
stable mutualist communities does not (see Supplementary Information).
Additionally, when each component of \(M\) is interpreted as a unique
species and given a random intrinsic growth
rate\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}},
feasibility is not increased by \(Var(\gamma)\), suggesting that
variation in species generation time might be unlikely to drive
stability in purely multi-species networks (see Supplementary
Information).

My results show that complex systems are more likely to be stable when
the response rates of system components vary. These results are broadly
applicable to complex biological and social networks.

@@

\hypertarget{ref-Hamblin2013}{}
18. Hamblin, S. On the practical usage of genetic algorithms in ecology
and evolution. \emph{Methods in Ecology and Evolution} \textbf{4,}
184--194 (2013).

\hypertarget{ref-Allesina2011}{}
19. Allesina, S. \& Levine, J. M. A competitive network theory of
species diversity. \emph{Proceedings of the National Academy of Sciences
of the United States of America} \textbf{108,} 5638--5642 (2011).

\hypertarget{ref-Dougoud2018}{}
20. Dougoud, M., Vinckenbosch, L., Rohr, R., Bersier, L.-F. \& Mazza, C.
The feasibility of equilibria in large ecosystems: a primary but
neglected concept in the complexity-stability debate. \emph{PLOS
Computational Biology} \textbf{14,} e1005988 (2018).
